Testing for measurement invariance with respect to an ordinal variable

### 2012-08-29 ordinal/ordinal.Rnw
Added fluctuation plots to application.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@166 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

## Remove cases with 'imputed' values
yg <- yg[yg$ao_4 <= 9,]
decimals <- apply(yg[,4:28]%%1 > 0, 1, sum)
yg <- yg[decimals==0,]

## TODO Simulate critical values for ordL2BB
@

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=1:4)
if(file.exists("ol2bb1.rda")){
  load("ol2bb1.rda")
} else {
  ol2bb1.fun <- ordL2BB(gefp1)
  save(ol2bb1.fun, file="ol2bb1.rda")
}
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)

@@

@@

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb2.rda")){
  load("ol2bb2.rda")
} else {
  ol2bb2.fun <- ordL2BB(gefp2)
  save(ol2bb2.fun, file="ol2bb2.rda")
}
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)

@@

\paragraph{Choice of Test}
The results presented in this paper imply that the proposed
ordinal statistics may ``miss'' measurement invariance violations that
are not monotonic w.r.t.\ $V$, moreso than the likelihood ratio test.  We
speculate that, in most applications, this will not be a major issue
because the researcher's a priori hypotheses exclusively focus on
monotonic measurement 
invariance violations.  For example, in the youth gratitude
application, we tested for measurement invariance across six age
groups.  If we observed a measurement invariance violation whereby
factor loadings were equal at all age groups except 14 years,
we would
have a hard time explaining the violation as anything but an anomaly
in the 14-year-olds.  Further, if $n$ is large, we are likely to
suspect that the result arises from the large sample size.
We may still be interested in why the
14-year-olds differed, but the analysis is purely exploratory at this
point because this type of violation was unexpected.  
However, there is generally a tradeoff between the ordinal statistics
and the likelihood ratio statistic.  The
proposed ordinal statistics usually provide sharper tests of one's a
priori hypothesis regarding measurement invariance w.r.t.\ ordinal
$V$, while the likelihood ratio statistic provides a sharper test of
general (non-monotonic) measurement invariance w.r.t.\ $V$.  While the
latter feature may be important in some high-stakes applications, many
researchers are likely to find the former feature appealing for their
work.

\paragraph{Extension to Other Models}
We focused on testing for measurement invariance in factor
analysis models here, but the proposed test statistics are applicable to
other psychometric models that are estimated via ML.  The only
requirement for carrying out the tests is that the casewise scores
(Equation~\eqref{eq:score}) be available following model estimation.
As a result, applications to studying DIF in IRT are immediate,
as are general psychometric applications to studying parameter
stability w.r.t.\ ordinal auxiliary variables.  We expect the
same general results to hold for these applications, whereby the
proposed test statistics are better than the LRT for detecting
monotonic instabilities.   The strucchange package \cite{Zei06} noted
previously can be used for these more-general applications.

\subsection{Summary}
The test statistics proposed in this paper have high power for
detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Bent80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous 
(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
advantageous properties that should be useful in practice.

>>>>

## Remove cases with 'imputed' values
yg <- yg[yg$ao_4 <= 9,]
decimals <- apply(yg[,4:28]%%1 > 0, 1, sum)
yg <- yg[decimals==0,]
@

@@

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, 
              order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=1:4)
if(file.exists("ol2bb1.rda")){
  load("ol2bb1.rda")
} else {
  ol2bb1.fun <- ordL2BB(gefp1)
  save(ol2bb1.fun, file="ol2bb1.rda")
}
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ol2bb1.fun)

@@

\begin{figure}
\caption{Fluctuation processes for the $\text{WDM}_o$ statistic (left panel) and
  the $\text{LM}_o$ statistic (right panel).}
\label{fig:gq6res}
\setkeys{Gin}{width=\textwidth}
<<gq6res, fig=TRUE, height=3, width=6>>=
par(mfrow=c(1,2), cex=.65)

plot(gefp1, functional=ordwmax(gefp1), axes=FALSE, main="",
     xlab="Age group", ylab=expression("WDM"[o]))
axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
axis(2)

plot(gefp1, functional=ol2bb1.fun, axes=FALSE, ylim=c(0,14), main="",
     xlab="Age group", ylab=expression("LM"[o]))
axis(1, at=1:5, labels=c("10-11","12-13","14","15","16"))
axis(2)
@
\end{figure}

@@

Plots representing the statistics' fluctuations across levels of age
group are displayed in Figure~\ref{fig:gq6res}.  The left panel
displays the process associated with $\text{WDM}_o$, while the right
panel displays the process associated with $\text{LM}_o$.  The final
age group (17-19 years) is not displayed, because the statistics
associated with this final age group are always zero (scores are
summed across all individuals in the sample at this point).
%% more detail

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, 
              order.by = as.numeric(yg$agegroup), 
              vcov = info_full, sandwich = FALSE, parm=3:6)
if(file.exists("ol2bb2.rda")){
  load("ol2bb2.rda")
} else {
  ol2bb2.fun <- ordL2BB(gefp2)
  save(ol2bb2.fun, file="ol2bb2.rda")
}
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ol2bb2.fun)

@@

\subsection{Choice of Test}
The results presented in this paper imply that the proposed
ordinal statistics may ``miss'' measurement invariance violations that
are not monotonic w.r.t.\ $V$, moreso than the likelihood ratio test.  We
speculate that, in most applications, this will not be a major issue
because the researcher's a priori hypotheses exclusively focus on
monotonic measurement 
invariance violations.  For example, in the youth gratitude
application, we tested for measurement invariance across six age
groups.  If we observed a measurement invariance violation whereby
factor loadings were equal at all age groups except 14 years,
we would
have a hard time explaining the violation as anything but an anomaly
in the 14-year-olds.  Further, if $n$ is large, we are likely to
suspect that the result arises from the large sample size.
We may still be interested in why the
14-year-olds differed, but the analysis is purely exploratory at this
point because this type of violation was unexpected.  
However, there is generally a tradeoff between the ordinal statistics
and the likelihood ratio statistic.  The
proposed ordinal statistics usually provide sharper tests of one's a
priori hypothesis regarding measurement invariance w.r.t.\ ordinal
$V$, while the likelihood ratio statistic provides a sharper test of
general (non-monotonic) measurement invariance w.r.t.\ $V$.  While the
latter feature may be important in some high-stakes applications, many
researchers are likely to find the former feature appealing for their
work.

\subsection{Extension to Other Models}
We focused on testing for measurement invariance in factor
analysis models here, but the proposed test statistics are applicable to
other psychometric models that are estimated via ML.  The only
requirement for carrying out the tests is that the casewise scores
(Equation~\eqref{eq:score}) be available following model estimation.
As a result, applications to studying DIF in IRT are immediate,
as are general psychometric applications to studying parameter
stability w.r.t.\ ordinal auxiliary variables.  We expect the
same general results to hold for these applications, whereby the
proposed test statistics are better than the LRT for detecting
monotonic instabilities.   The strucchange package \cite{Zei06} noted
previously can be used for these more-general applications.

\subsection{Summary}
The test statistics proposed in this paper have relatively-high power for
detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have relatively-low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous 
(e.g., \citeNP{BauHus09,MolDol10,NeaAgg06}).  In all, the tests have
advantageous properties that should be useful in practice.
